Higher-Order unification for free!

### 2024-04-11 paper/main.tex
wip: maybebeta section

\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
\end{corollary}

@@

\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
  not a name.
\end{corollary}

@@

\paragraph{Decompilation}
During \elpiIn{progress}, as claimed in \fullref{inv:beta-rhs}, the decompilation
can only have \linkbeta with not instantiated \lhs. In this case, \lhs is
unified with \rhs.
\todo{qui falliamo, nella 7.3 dire che LHS è in llambda quindi LSH = Ugly può funzionare,
in Elpi funziona però in questi casi il termine in llam deve essere a sinistra}.

@@

\newcommand{\progBetaLL}{\emph{progress-\llambda-refine}\xspace}
\newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
\newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
\newcommand{\progBetaLH}{\emph{progress-\llambda-\lhs}\xspace}

We give $4$ different definitions of \linkbeta progression for a given \linkbeta
\linkbetaM{\Gamma}{T}{\rhsBeta}, say $l$.

@@

% Example of this `X a = Y a'
\begin{definition}[\progBetaDedup]
  Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
  share the same \lhs. The two \rhs are unified and $l_2$ is
  removed from \linkStore.
  \label{def:progBetaDedup}
\end{definition}

\begin{definition}[\progBetaLH]
  Given a \linkbeta with rigid \lhs, the unification fails.
  \label{def:progBetaLH}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaDedup (resp. \progBetaLH) also guarantees termination since 
  it removes a link from \linkStore (resp. immediately fails).

@@

\begin{lemma}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity}
\end{lemma}

@@

\subsection{Relaxing \fullref{def:progBetaLH}}

@@

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive. There exists systems
such as $\lambda$Prolog~\cite{lamProlog}, Abella~\cite{gacek2008abella}, which 
delay the resolution of \notllambda unification problems if the substitution
is not able to put them in \llambda.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.Y}}}
  {{}}
  {{}}
  {{}}

>>>>

\noteD{Si puo togliere se serve spazio}
\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
\end{corollary}

@@

\noteD{Si puo togliere se serve spazio}
\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
  not a name.
\end{corollary}

@@

\paragraph{Decompilation}
We suppose that the unification of the object language rejects any unification
between two terms where at least one of them is not in \llambda 
(this includes unifications like $X = Y\appsep Z$ and $X\appsep Y = X \appsep Y$).
Therefore, a failure is thrown if any \linkbeta remains in \linkStore at the
begin of decompilation, i.e. all \linkbeta should be solved before decompilation.

@@

\newcommand{\progBetaLL}{\emph{progress-\llambda-refine}\xspace}
\newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
% \newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
\newcommand{\progBetaFail}{\emph{progress-\llambda-fail}\xspace}

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
$4$ different activation rules:

@@

% Example of this `X a = Y a'
% \begin{definition}[\progBetaDedup]
%   Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
%   share the same \lhs. The two \rhs are unified and $l_2$ is
%   removed from \linkStore.
%   \label{def:progBetaDedup}
% \end{definition}

\begin{definition}[\progBetaFail]
  If it exists a link $l' \in \linkStore$ with same \lhs as $l$, or the \lhs 
  of $l$ become rigid, then unification fail.
  \footnote{This is a consequence of what claimed in \linkbeta decompilation}
  \label{def:progBetaFail}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity}
\end{theorem}

@@

\subsection{Relaxing \fullref{def:progBetaFail}}

@@

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
There exists systems
such as $\lambda$Prolog~\cite{lamProlog} which 
delay the resolution of \notllambda unification problems if the substitution
is not able to put them in \llambda. For example in 
\noteD{Teijus}
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.Y}}}
  {{}}
  {{}}
  {{}}
